---
title: "3rd-Assignment"
author: "Andrea Češková, Tobias Grüner, Tuan Linh Tran"
format: html
editor: visual
---

```{r set-up, include=FALSE}
library(pacman)

pacman::p_load(
  httr,
 jsonlite,
  ggplot2,
 keyring,
 ecmwfr,
  dplyr,
reticulate,
  geodata,
  sf,
 rvest,
  raster,
  rnaturalearth,
 rnaturalearthdata,
 grid,
 readr,
 ggspatial,
 gridExtra,
 kableExtra,
 ggrepel,
 tidygeocoder,
 osrm ,
  here,
  terra,
  tidyr,
  patchwork
)
here::i_am("R/3rd-Assignment.qmd")
```

# Introduction

We examine the correlation between population density and temperature across different land use types in the Netherlands. Our goal is to understand how land use mediates the relationship between temperature and population.

# Data

## Administrative boundaries

```{r}
nld_2 <- st_read(here("data", "raw", "gadm41_NLD_shp", "gadm41_NLD_2.shp"))
```

## Population Density

```{r}
pop <- raster(here("data","raw", "population_nld_2019-07-01_geotiff", "population_nld_2019-07-01.tif"))
```

## Temperature

(i dont know how to download the temp data from chelsa, so i just tried other source)

```{r API-process}

# set API key 
key_set(
  service  = "cds_api",
  username = "linhtran01091310@gmail.com"
)

# now pull it back out:
cds_api_key <- key_get(
  service  = "cds",
  username = "linhtran01091310@gmail.com"
)

# Export to environment variables for cdsapi, reticulate, or ecmwfr
Sys.setenv(
  CDSAPI_URL = "https://cds.climate.copernicus.eu/api",
  CDSAPI_KEY = cds_api_key
)
```

```{r}
wf_set_key(
  key  = cds_api_key
)

wf_get_key()
request <- list(
  dataset_short_name  = "reanalysis-era5-land",
  product_type        = "reanalysis",
  variable            = "2m_temperature",
  year                = "2019",
  month               = "07",
  day                 = sprintf("%02d", 1:31),         # all July days
  time                = sprintf("%02d:00", 0:23),       # hourly
  area                = c(53.6, 3.2, 50.7, 7.2),        # N, W, S, E (Netherlands)
  data_format         = "netcdf",                      # better for R
  download_format     = "unarchived",
  target              = "netherlands_july2019.nc"      # output filename
)

era5 <- wf_request(
  request  = request,
  transfer = TRUE,
  path     = getwd()
)
```

```{r}
temp <- rast("D:/GIS assignment 3/GIS-3rd-assignment/R/netherlands_july2019.nc")
```

## Land Use

```{r}

```
